---
---

## Characteristics of a Database: Part I

Let's begin with defining some terminology that is familiar from any system for data storage.
For concreteness, we will use the case study just introduced to provide examples.

Field
: The smallest unit of information, each having a label and holding a value of the same type.
: e.g. The time of the survey in which a small mammal was captured.

Record
: A collection of related values, from different fields, that all describe the same entity.
: e.g. The species, sex, size and location of a small mammal observed during a given survey.
  
Table
: A collection of records, each one uniquely identified by the value of one or more fields.
: e.g. All records of small mammals that were observed in any survey.

These characteristics could equally be applied to a spreadsheet, one that can be curated in a text file (e.g. a CSV file) as well as specialized formats (e.g. a MS Excel file).
{:.notes}

===

## Limitations of Data Files

Collaboration
: {:.fragment} Email copies among collaborators, store in the cloud (sync issues) or save to a network (user collision).

Size
: {:.fragment} Reading an entire data file into memory isn't scaleable. Some file types (e.g. MS Excel files) have size limits.

Quality
: {:.fragment} Enforcing restrictions on values is not the default behaviour of software that writes data files.

Extensions
: {:.fragment} Specialized files are needed for non-traditional data types (e.g. ESRI Shapefiles).

API
: {:.fragment} No standard approach for different application to read, edit or create records.

===

## Database Solutions

Collaboration
: {:.fragment} A database accepts simultaneous users, which is most beneficial when the database is hosted on a network. There are never multiple copies of the data (aside from your backups!)

Size
: {:.fragment} The database management system only reads requested parts of the data into memory. There is no size limit!

Quality
: {:.fragment} Data types are enforced by default. No database lets you enter "9I1" in an integer field.

Extension
: {:.fragment} Special packages are needed for non-traditional data types, but they can live in the same table as traditional ones.

API
: {:.fragment} There are packages native to every programming language that ease reading and writing to databases.

===

## Database connections from R

```{r message=FALSE, title="lesson-3.R", eval=FALSE}
library(RPostgreSQL)
con <- dbConnect(PostgreSQL(), host="postgresql.sesync.org", dbname="portal")
```

===

## Database connections from R

```{r message=FALSE, title="lesson-3.R"}
library(RSQLite)
con <- dbConnect(SQLite(), "data/portal.sqlite")
```

```{r}
dbListTables(con)
```

===

## Two ways to access data

Read the data into an R data frame, and process it using R functions.

```{r title="lesson-3.R"}
plots <- dbReadTable(con, "plots")
surveys <- dbReadTable(con, "surveys")
species <- dbReadTable(con, "species")
```

```{r}
str(surveys)
```

===

Let the database system process data.

```{r title="lesson-3.R"}
dbGetQuery(con, "select species_id, weight
                 from surveys
                 where plot_id = 1 limit 5")
```

The string inside the brackets is an example of Structured Query Language (SQL). The SQL instructions tell the database system (SQLite, in this case) to sort out and return only the records requested.

===

## Exercise 1

Use `dbGetQuery()` to select the "species_id", and two other fields from the "species" table. Hint: use `dbListFields()` to check field names.
B